---
title: |
  <center> Incomplete Data Analysis Assignment Code Markdown </center>
author: "Jackson Cramer, Maximillian McCourt, Brandon Causing"
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document:
    df_print: paged
  pdf_document: default
---


Group Member Contributions:<br>
- Brandon Causing (s2901457): Coding and write-up of Task 1 (roughly 1/3 of total work).<br>
- Jackson Cramer (s2274544): Coding and report write-up of Task 2, questions 1-3. (roughly 1/3 of total work).<br>
- Maximillian McCourt (s2145762): Coding and report write-up of Task 2, questions 4-5, as well as final edits (roughly 1/3 of total work).<br>


```{r}
library(VIM)
library(naniar)
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(mice)
library(MASS)

```


# Task 1. Comparison of complete case analysis, stochastic regression imputation and predictive mean matching

## 1.

```{r}
data(birthwt)
```

```{r}
# make sure data is complete
any(is.na(birthwt$age))
any(is.na(birthwt$bwt))
```


```{r}
# set up complete dataset (age vs. birth weight)
x <- birthwt$age
y <- birthwt$bwt

data_complete <- data.frame(x = x, y = y)
```

```{r}
# simulate missingess
set.seed(1)

y_scaled <- scale(y) # scale for stability
p_miss <- plogis(0.5 * y_scaled) # logistic model for missingness prob
R <- rbinom(length(x), 1, p_miss) # generate missingness (1 -> missing)

# substitute the missing values into the dataset
x_obs <- x
x_obs[R == 1] <- NA

data_obs <- data.frame(x = x_obs, y = y)
```

```{r}
# about 50% of the data should be missing
mean(is.na(data_obs$x))
```

## 2.

```{r}
# fit x on y regression for both the original and the missing data datasets
fit_full_xy <- lm(x ~ y, data = data_complete)
fit_missing_xy <- lm(x ~ y, data = data_obs)

# summary outputs
summary(fit_full_xy)
summary(fit_missing_xy)
```

```{r}
data_complete %>% ggplot(aes(x = y, y = x)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) +
  labs(title = "Complete Data (x = 21.3 + 0.00066y)", x = "Birth Weight (g)", y = "Maternal Age") + 
  theme_bw()

data_obs %>% ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  labs(title = "Missing Data (x = 24.4 - 0.0007y)", x = "Birth Weight (g)", y = "Maternal Age") + 
  theme_bw()
```

```{r}
# residual plots, nothing too alarming
par(mfrow = c(2,2))
plot(fit_missing_xy)
plot(fit_full_xy)
par(mfrow = c(1,1))
```

## 3.

```{r}
# fit y on x regression for both the original and the missing data datasets
fit_full_yx <- lm(y ~ x, data = data_complete)
fit_missing_yx <- lm(y ~ x, data = data_obs)

# summary outputs
summary(fit_full_yx)
summary(fit_missing_yx)
```

```{r}
data_complete %>% ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) +
  labs(title = "Complete Data (y = 2656 + 12.43x)", x = "Maternal Age", y = "Birth Weight (g)") + 
  theme_bw()

data_obs %>% ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  labs(title = "Missing Data (y = 3084 - 15.48x)", x = "Maternal Age", y = "Birth Weight (g)") + 
  theme_bw()
```

```{r}
# residual plots, nothing too alarming
par(mfrow = c(2,2))
plot(fit_missing_yx)
plot(fit_full_yx)
par(mfrow = c(1,1))
```

## 4.

```{r}
# fit x on y on missing data dataset
fit_imp_xy <- lm(x ~ y, data = data_obs)
```

```{r}
# predicted values
x_pred <- predict(fit_imp_xy, newdata = data_obs)

# residual st. dev.
sigma <- summary(fit_imp_xy)$sigma

# find indices of missing observations
x_imp <- data_obs$x
missings <- is.na(x_imp)

# replace NAs with predicted + random noise
x_imp[missings] <- x_pred[missings] + rnorm(sum(missings), 0, sigma)

data_imp <- data.frame(x = x_imp, y = y, missing = as.factor(missings*1))
```

```{r}
# fit y on x using data with imputed values
fit_imp_yx <- lm(y ~ x, data = data_imp) 
summary(fit_imp_yx)
```

```{r}
data_imp %>% ggplot(aes(x = x, y = y)) + 
  geom_point(aes(color = missing)) + 
  geom_smooth(method = 'lm', se = FALSE, color = 'purple') + 
  labs(title = "Stochastic Regression Imputation (y = 3226 - 12.65x)",
       x = "Maternal Age", y = "Birth Weight (g)" ) + 
  scale_color_discrete(name = "Missing") + 
  theme_bw()
```

## 5.

```{r}
# 100 simulations
n_sim <- 100

# initialize list of results (int, slope, var, corr)
results <- data.frame(
  
  # complete-case
  intercept_cc = numeric(n_sim),
  slope_cc = numeric(n_sim),
  var_cc = numeric(n_sim),
  cor_cc = numeric(n_sim),
  
  # imputed
  intercept_imp = numeric(n_sim),
  slope_imp = numeric(n_sim),
  var_imp = numeric(n_sim),
  cor_imp = numeric(n_sim)
)
```

```{r}
# run the simulations
for (i in 1:n_sim) {
  
  # --- Complete-Case ---
  
  # generate missingness
  p_miss <- plogis(0.5 * scale(y))
  R <- rbinom(length(x), 1, p_miss)
  
  # substitute in missing values
  x_obs <- x
  x_obs[R == 1] <- NA
  data_obs <- data.frame(x = x_obs, y = y)
  
  # complete-case regression
  fit_cc <- lm(y ~ x, data = data_obs)
  
  # store results
  results$intercept_cc[i] <- coef(fit_cc)[1]
  results$slope_cc[i] <- coef(fit_cc)[2]
  results$var_cc[i] <- summary(fit_cc)$sigma^2
  results$cor_cc[i] <- cor(data_obs$x, data_obs$y, use = "complete.obs")
  
  
  # --- Stochastic Regression Imputation ---
  
  # fit x on y
  fit_imp_xy <- lm(x ~ y, data = data_obs)
  
  # predicted values + residual st. dev.
  x_pred <- predict(fit_imp_xy, newdata = data_obs)
  sigma <- summary(fit_imp_xy)$sigma
  
  # replace NAs with predicted + random noise
  x_imp <- x_obs
  missings <- is.na(x_imp)
  x_imp[missings] <- x_pred[missings] + rnorm(sum(missings), 0, sigma)
  
  data_imp <- data.frame(x = x_imp, y = y)
  
  # fit y on x using data with imputed values
  fit_imp_yx <- lm(y ~ x, data = data_imp)
  
  # store results
  results$intercept_imp[i] <- coef(fit_imp_yx)[1]
  results$slope_imp[i] <- coef(fit_imp_yx)[2]
  results$var_imp[i] <- summary(fit_imp_yx)$sigma^2
  results$cor_imp[i] <- cor(data_imp$x, data_imp$y)
}
```

```{r}
# summary statistics of complete dataset for comparison
complete_model <- lm(y ~ x, data = data_complete)

complete_intercept <- coef(complete_model)[1]
complete_slope <- coef(complete_model)[2]
complete_var <- summary(complete_model)$sigma^2
complete_cor <- cor(x, y)
```

```{r}
# slope histograms
results %>% ggplot(aes(x = slope_cc)) +
  geom_histogram(binwidth = 5, fill = 'lightblue') + 
  geom_vline(xintercept = complete_slope, color = 'red', linewidth = 1) + 
  geom_vline(xintercept = mean(results$slope_cc), color = 'blue', linewidth = 1) + 
  labs(title = "Complete-Case Analysis (avg = 1.04, obs = 12.43)", x = "Age Coefficient",
       y = NULL) + 
  theme_bw()

results %>% ggplot(aes(x = slope_imp)) +
  geom_histogram(binwidth = 5, fill = 'lightblue') + 
  geom_vline(xintercept = complete_slope, color = 'red', linewidth = 1) + 
  geom_vline(xintercept = mean(results$slope_imp), color = 'blue', linewidth = 1) +
  labs(title = "Stochastic Regression Imputation (avg = 0.23, obs = 12.43)", x = "Age Coefficient",
       y = NULL) + 
  theme_bw()
```

```{r}
# intercept histograms
results %>% ggplot(aes(x = intercept_cc)) +
  geom_histogram(bins = 10, fill = 'lightblue') + 
  geom_vline(xintercept = complete_intercept, color = 'red', linewidth = 1) + 
  geom_vline(xintercept = mean(results$intercept_cc), color = 'blue', 
             linewidth = 1) +
  labs(title = "Complete-Case Analysis (avg = 2749.9, obs = 2655.7)", x = "Intercept Coefficient",
       y = NULL) + 
  theme_bw()

results %>% ggplot(aes(x = intercept_imp)) +
  geom_histogram(bins = 10, fill = 'lightblue') + 
  geom_vline(xintercept = complete_intercept, color = 'red', linewidth = 1) + 
  geom_vline(xintercept = mean(results$intercept_imp), color = 'blue', 
             linewidth = 1) + 
  labs(title = "Stochastic Regression Imputation (avg = 2937.7, obs = 2655.7)", x = "Intercept Coefficient",
       y = NULL) + 
  theme_bw()
```

```{r}
# variance histograms
results %>% ggplot(aes(x = var_cc)) +
  geom_histogram(bins = 10, fill = 'lightblue') + 
  geom_vline(xintercept = complete_var, color = 'red', linewidth = 1) + 
  geom_vline(xintercept = mean(results$var_cc), color = 'blue', linewidth = 1) + 
  labs(title = "Complete-Case Analysis (avg = 512803.1, obs = 530236.2)", x = "Residual Variance",
       y = NULL) + 
  theme_bw()

results %>% ggplot(aes(x = var_imp)) +
  geom_histogram(bins = 10, fill = 'lightblue') + 
  geom_vline(xintercept = complete_var, color = 'red', linewidth = 1) + 
  geom_vline(xintercept = mean(results$var_imp), color = 'blue', linewidth = 1) + 
  labs(title = "Stochastic Regression Imputation (avg = 531066.9, obs = 530236.2)", x = "Residual Variance",
       y = NULL) + 
  theme_bw()
```

```{r}
# correlation histograms
results %>% ggplot(aes(x = cor_cc)) +
  geom_histogram(bins = 10, fill = 'lightblue') + 
  geom_vline(xintercept = complete_cor, color = 'red', linewidth = 1) +
  geom_vline(xintercept = mean(results$cor_cc), color = 'blue', linewidth = 1) +
  labs(title = "Complete-Case Analysis (avg = 0.009, obs = 0.090)", x = "Correlation Coefficient",
       y = NULL) + 
  theme_bw()

results %>% ggplot(aes(x = cor_imp)) +
  geom_histogram(bins = 10, fill = 'lightblue') + 
  geom_vline(xintercept = complete_cor, color = 'red', linewidth = 1) + 
  geom_vline(xintercept = mean(results$cor_imp), color = 'blue', linewidth = 1) +
  labs(title = "Stochastic Regression Imputation (avg = 0.003, obs = 0.090)", x = "Correlation Coefficient",
       y = NULL) + 
  theme_bw()
```


# Task 2. Analysis of NHANES data regarding risk factors of cardiovascular disease

## 1.


```{r}
## Assignment IDA 25/26 reading in NHANES data

# set the working directory to the location where the data file is

dat<-dget(file="NHANESvIDA.rda")
str(dat) # concise info on size and variables
dat$id <- NULL # delete observation id as not required
head(dat) # first six observations
summary(dat) # summary for all variables 

```

## 2.


```{r}
# Aggregation plot of missing data patterns

# create aggregation object
v <- aggr(dat, plot=FALSE)

# plot missingness patterns (counts shown instead of proportions)
plot(v, prop=FALSE, numbers=TRUE)

```


```{r}
# missingness stratified by categorical variables

# by gender (exclude other categorical vars to avoid redundancy)

gg_miss_var(
  dat %>% dplyr::select(-marital, -ethnic, -educ, -income, -num_mvpa_bouts, -cpm, -age),
  facet = gender,
  show_pct = TRUE
) +
  ggtitle("Gender") +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

# by ethnicity
gg_miss_var(
  dat %>% dplyr::select(-gender, -marital, -educ, -income, -num_mvpa_bouts, -cpm, -age),
  facet = ethnic,
  show_pct = TRUE
) +
  ggtitle("Ethnicity") +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

# by education level
gg_miss_var(
  dat %>% dplyr::select(-gender, -marital, -ethnic, -income, -num_mvpa_bouts, -cpm, -age),
  facet = educ,
  show_pct = TRUE
) +
  ggtitle("Education") +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

# by marital status
gg_miss_var(
  dat %>% dplyr::select(-gender, -educ, -ethnic, -income, -num_mvpa_bouts, -cpm),
  facet = marital,
  show_pct = TRUE
) +
  ggtitle("Marital Status") +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

# remove missing income values before faceting (since it's the stratifying variable)
dat_no_income_na <- dat %>% filter(!is.na(income))

# by income
gg_miss_var(
  dat_no_income_na %>% dplyr::select(-gender, -marital, -educ, -ethnic, -num_mvpa_bouts, -cpm, -age),
  facet = income,
  show_pct = TRUE
) +
  ggtitle("Income") +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

```



```{r}
# margin plots: relationship between predictors and LDL-cholesterol (with missingness shown)

# height vs cholesterol
marginplot(dat[c(10, 7)],
           main = "Height vs. Cholesterol",
           xlab = "Height (cm)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

# BMI vs cholesterol
marginplot(dat[c(10, 8)],
           main = "BMI vs. Cholesterol",
           xlab = "Body mass index (kg/m^2)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

# waist circumference vs cholesterol
marginplot(dat[c(10, 9)],
           main = "Waist vs. Cholesterol",
           xlab = "Waist circumference (cm)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

# apolipoprotein B vs cholesterol
marginplot(dat[c(10, 11)],
           main = "Apoliprotein B vs. Cholesterol",
           xlab = "Apoliprotein (B) (mg/dL)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

# mean activity counts vs cholesterol
marginplot(dat[c(10, 12)],
           main = "Mean counts vs. Cholesterol",
           xlab = "Means counts per day",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

# MVPA bouts vs cholesterol
marginplot(dat[c(10, 13)],
           main = "Mean MVPA bouts vs. Cholesterol",
           xlab = "Mean MVPA bouts per day",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

```


## 3.

```{r}
# univariate summaries of cholesterol

# numerical summary
print(summary(dat$cholesterol))

# remove missing values for plotting
non_na_cholesterol <- dat[!is.na(dat$cholesterol), ]

par(mfrow = c(1, 2))

# box plot of cholesterol
boxplot(non_na_cholesterol$cholesterol,
        main = "Boxplot of Cholesterol",
        ylab = "LDL-Cholesterol (mg/dL)",
        col = "lightblue",
        border = "steelblue")

# density plot of cholesterol
dens <- density(non_na_cholesterol$cholesterol)
plot(dens,
     main = "Density of Cholesterol",
     xlab = "LDL-Cholesterol (mg/dL)",
     col = "darkorange",
     lwd = 2)
polygon(dens, col = "moccasin", border = "darkorange")

par(mfrow = c(1, 1))

```


```{r}
# bivariate summaries of cholesterol

# filter out missing income values for plotting
non_na_cholesterol <- non_na_cholesterol %>%
  filter(!is.na(income))

# pivot categorical variables to long format for box plots
dat_long <- non_na_cholesterol %>%
  pivot_longer(
    cols =  c(marital, ethnic, educ, income, gender),
    names_to = "variable",
    values_to = "category"
  )

# boxplots of cholesterol by categorical variable
ggplot(dat_long, aes(x = category, y = cholesterol)) +
  geom_boxplot(fill = "skyblue") +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  labs(
    x = "",
    y = "LDL-Cholesterol (mg/dL)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


# pivot numeric predictors to long format for scatterplots
dat_long_num <- non_na_cholesterol %>%
  pivot_longer(
    cols = c(age, height, bmi, waist, cpm, num_mvpa_bouts),
    names_to = "variable",
    values_to = "value"
  )

# scatterplots of cholesterol vs numeric predictors
ggplot(dat_long_num, aes(x = value, y = cholesterol)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  labs(
    x = "Predictor Value",
    y = "LDL-Cholesterol (mg/dL)"
  )

# pearson correlation matrix for numeric variables
num_data <- dat %>%
  dplyr::select(cholesterol, age, height, bmi, waist, cpm, num_mvpa_bouts) %>%
  na.omit()
cor_mat <- cor(num_data, method = "pearson")

# plot correlation matrix
corrplot(cor_mat,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         number.cex = 0.7)
```

## 4.

Let's set up our linear model

```{r}
# add indicator for mvpa 
dat['ind_mvpa'] <- as.numeric(dat['num_mvpa_bouts'] != 0)

# fit linear model with cholesterol as response variable
m1 <- lm(cholesterol ~ gender + age + ethnic + educ + marital + income + height + waist
    + cpm + ind_mvpa, data=dat)

summary(m1)
```

We next investigate the validity of our model assumptions

```{r}

par(mfrow=c(2,2))
plot(m1)
ggsave("modelfit.png", width=8, height=5, dpi=300)

```


## 5.


```{r}

# let's first set up the complete case dataset and model fit

# Complete Case Analysis
dat_comp <- na.omit(dat)
dat_comp$apoliprot <- NULL

# complete case analysis model fit
fit_cc <- lm(cholesterol ~ gender + age + ethnic + educ + marital + income + 
               height + waist + cpm + ind_mvpa, data=dat_comp)

# summary statistics
summary(fit_cc)

```


Initialise dataset for modelling

```{r}
# install.packages("mice")
md.pattern(dat)

# Subset to only data where cholesterol is observed
dat_nona <- dat[!is.na(dat$cholesterol), ]

# Re-create ind_mvpa on this subset for imputation setup
dat_nona$ind_mvpa <- as.numeric(dat_nona$num_mvpa_bouts != 0)

# Drop num_mvpa_bouts as already encoded it with ind_mvpa, 
dat_nona$num_mvpa_bouts <- NULL
dat_nona$apoliprot <- NULL # also drop apoliprot

# extract names of variables with missing data
missing_var <- names(dat_nona)[apply(is.na(dat_nona), 2, any)]

# find rows with missing values in dat_nona
missing_idx <- rowSums(is.na(dat_nona))
cat("Total rows with at least one missing value:", sum(missing_idx > 0), "\n")

# View those rows
dat_nona[missing_idx > 0, ]

# Also useful: which specific variables are missing for each row
colSums(is.na(dat_nona))

```


Do an initial dry run using mice() function

```{r}
# dry run to obtain the default methods and predictor matrix
setup <- mice(dat_nona, maxit=0, seed=1)

pred <- setup$predictorMatrix
meth <- setup$method
# do not impute response (cautionary measure)
meth["cholesterol"] <- ""
# display methods
meth

# deal with potential multicollinearity between bmi and waist
# by removing bmi as a imputing predictor of waist and vice versaa
pred['bmi', 'waist'] <- 0
pred['waist', 'bmi'] <- 0
#cor(dat_nona$waist, dat_nona$bmi)

# restrict bmi values
post <- setup$post
post["bmi"] <- "imp[[j]][,i] <- squeeze(imp[[j]][,i], c(10, 60))"


# dry run (30 imputations, 20 iterations)
dat_mice <- mice(dat_nona, method = meth, predictorMatrix = pred, post=post,#visitSequence = visSeq,
                  maxit = 20, m = 30, seed = 1, printFlag = FALSE)

# Check for any logged problems during imputation
dat_mice$loggedEvents

```


```{r fig.height=4, fig.width=10, outheight="95%"}

# plot 
#par(mfrow=c(4,1))
plot(dat_mice, c("income", "height", "bmi", "waist"), layout=c(2,4))

# density plot
densityplot(dat_mice, 
            lwd = 2,
            col = c("steelblue", "indianred"),  # blue = observed, red = imputed
            scales = list(y = list(relation = "free"),
                         x = list(relation = "free")),
            xlab = "Value",
            #strip = strip.custom(
            #  bg = "grey90",
            #  par.strip.text = list(cex = 0.9, font = 2)
            #),
            par.settings = list(
              plot.line = list(lwd = 2),
              superpose.line = list(
                col = c("steelblue", "indianred"),
                lwd = 2
              ),
              background = list(col = "white"),
              strip.background = list(col = "grey90")
            ))

# visualise imputed values (grey: observed, red: imputed) for each variable
stripplot(dat_mice, height + waist + bmi ~ .imp,
          scales = list(x = list(at = c(0, 5, 10, 15, 20, 25, 30),
                                  labels = c("0","5","10","15","20","25","30")),
                        y = list(relation = "free")),
          col = c("grey", mdc(2)),
          pch=20, cex=1.2,
          xlab = "Imputation number (0 = observed)")

# visualise imputations made for BMI at each iteration
stripplot(dat_mice, bmi ~ .imp, pch = 20, cex = 1.2,
          xlab = "Imputation number (0 = observed)",
          main = "Observed and imputed values of BMI")

```


Fitting model

```{r}

# fit the model using mice-imputed data
fit <- with(dat_mice, lm(cholesterol ~ gender + age + ethnic + educ + marital + 
                            income + height + waist + cpm + ind_mvpa))

# pool data and find summary statistics
pool_fit <- pool(fit)
summary(pool_fit, conf.int=TRUE)

# explained variance of our pooled fitted data
pool.r.squared(pool_fit, adjusted=TRUE)

```










