---
title: "maxstuff"
output: html_document
date: "2026-03-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import data



```{r}
## Assignment IDA 25/26 reading in NHANES data
library(ggplot2)

# set the working directory to the location where the data file is

dat<-dget(file="NHANESvIDA.rda")
str(dat) # concise info on size and variables
dat$id <- NULL # delete observation id as not required
head(dat) # first six observations
summary(dat) # summary for all variables 


```

## (4) Linear modelling

Let's set up our linear model

```{r}
head(dat)

# add indicator for mvpa 
dat['ind_mvpa'] <- as.numeric(dat['num_mvpa_bouts'] != 0)

# fit linear model with cholesterol as response variable
m1 <- lm(cholesterol ~ gender + age + ethnic + educ + marital + income + height + waist
    + cpm + ind_mvpa, data=dat)

summary(m1)
```

We next investigate the validity of our model assumptions

```{r}

par(mfrow=c(2,2))
plot(m1)
ggsave("modelfit.png", width=8, height=5, dpi=300)

```


-Residuals vs Fitted: Red indicator line is relatively flat, centred near zero -> linearity assumption holds.
-Q-Q Residuals: Residuals do not follow normal distribution well, lower quantiles especially -> some skewness.
-Scale-Location: Clear signs of heteroskedasticity -> non-constant variance...
-Residuals vs Leverage: Outliers do not appear to violate distribution too much (Cook's distance < 0.5).


## (5) Missing values

Dealing with missing values. Regarding the response variable of your choice use cases
where the response is not missing, that is do not impute the response! Regarding the predictor variables choose a single imputation method (one of the methods we covered including predictive mean matching) or multiple imputation. Justify your choice based on the
missing value mechanism you are assuming (MCAR, MAR or MNAR) and the merits of the
method. Also carry out an analysis which is entirely based on complete cases. Comment
on the difference in results between complete case analysis and the analysis based on some
form of imputation. How do the results differ and are there big differences? What are the
advantages and/or disadvantages of the analysis based on imputation over the complete
case analysis? Justify your answer


```{r}

# let's first set up the complete case dataset and model fit

# Complete Case Analysis
dat_comp <- na.omit(dat)
dat_comp$apoliprot <- NULL

# complete case analysis model fit
fit_cc <- lm(cholesterol ~ gender + age + ethnic + educ + marital + income + 
               height + waist + cpm + ind_mvpa, data=dat_comp)

# summary statistics
summary(fit_cc)

```


Initialise dataset for modelling

```{r}
# install.packages("mice")
library(mice)
md.pattern(dat)

# Subset to only data where cholesterol is observed
dat_nona <- dat[!is.na(dat$cholesterol), ]

# Re-create ind_mvpa on this subset for imputation setup
dat_nona$ind_mvpa <- as.numeric(dat_nona$num_mvpa_bouts != 0)

# Drop num_mvpa_bouts as already encoded it with ind_mvpa, 
dat_nona$num_mvpa_bouts <- NULL
dat_nona$apoliprot <- NULL # also drop apoliprot

# extract names of variables with missing data
missing_var <- names(dat_nona)[apply(is.na(dat_nona), 2, any)]

# find rows with missing values in dat_nona
missing_idx <- rowSums(is.na(dat_nona))
cat("Total rows with at least one missing value:", sum(missing_idx > 0), "\n")

# View those rows
dat_nona[missing_idx > 0, ]

# Also useful: which specific variables are missing for each row
colSums(is.na(dat_nona))

```


Do an initial dry run using mice() function

```{r}
# dry run to obtain the default methods and predictor matrix
setup <- mice(dat_nona, maxit=0, seed=1)

pred <- setup$predictorMatrix
meth <- setup$method
# do not impute response (cautionary measure)
meth["cholesterol"] <- ""
# display methods
meth

# deal with potential multicollinearity between bmi and waist
# by removing bmi as a imputing predictor of waist and vice versaa
pred['bmi', 'waist'] <- 0
pred['waist', 'bmi'] <- 0
#cor(dat_nona$waist, dat_nona$bmi)

# restrict bmi values
post <- setup$post
post["bmi"] <- "imp[[j]][,i] <- squeeze(imp[[j]][,i], c(10, 60))"


# dry run (30 imputations, 20 iterations)
dat_mice <- mice(dat_nona, method = meth, predictorMatrix = pred, post=post,#visitSequence = visSeq,
                  maxit = 20, m = 30, seed = 1, printFlag = FALSE)

# Check for any logged problems during imputation
dat_mice$loggedEvents

```


```{r fig.height=4, fig.width=10, outheight="95%"}

# plot 
#par(mfrow=c(4,1))
plot(dat_mice, c("income", "height", "bmi", "waist"), layout=c(2,4))

# density plot
densityplot(dat_mice, 
            lwd = 2,
            col = c("steelblue", "indianred"),  # blue = observed, red = imputed
            scales = list(y = list(relation = "free"),
                         x = list(relation = "free")),
            xlab = "Value",
            #strip = strip.custom(
            #  bg = "grey90",
            #  par.strip.text = list(cex = 0.9, font = 2)
            #),
            par.settings = list(
              plot.line = list(lwd = 2),
              superpose.line = list(
                col = c("steelblue", "indianred"),
                lwd = 2
              ),
              background = list(col = "white"),
              strip.background = list(col = "grey90")
            ))

# visualise imputed values (grey: observed, red: imputed) for each variable
stripplot(dat_mice, height + waist + bmi ~ .imp,
          scales = list(x = list(at = c(0, 5, 10, 15, 20, 25, 30),
                                  labels = c("0","5","10","15","20","25","30")),
                        y = list(relation = "free")),
          col = c("grey", mdc(2)),
          pch=20, cex=1.2,
          xlab = "Imputation number (0 = observed)")

# visualise imputations made for BMI at each iteration
stripplot(dat_mice, bmi ~ .imp, pch = 20, cex = 1.2,
          xlab = "Imputation number (0 = observed)",
          main = "Observed and imputed values of BMI")

```


Fitting model

```{r}

# fit the model using mice-imputed data
fit <- with(dat_mice, lm(cholesterol ~ gender + age + ethnic + educ + marital + 
                            income + height + waist + cpm + ind_mvpa))

# pool data and find summary statistics
pool_fit <- pool(fit)
summary(pool_fit, conf.int=TRUE)

# explained variance of our pooled fitted data
pool.r.squared(pool_fit, adjusted=TRUE)

```
































