---
title: |
  <center> Incomplete Data Analysis Assignment Code Markdown </center>
author: "Jackson Cramer, Maximillian McCourt, Brandon Causing"
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document:
    df_print: paged
  pdf_document: default
---


Group Member Contributions:<br>
- Brandon Causing - s2901457: Coding and write up of Task 1. (roughtly 1/3 of total work)<br>
- Maximillian McCourt - s2145762: Coding and report write up of Task 2, questions 1-3, as well as editing of the remaining writeup of Task 2. (roughly 1/3 of total work)<br>
- Jackson Cramer - s2274544: Coding and report write up of Task 2, questions 4-5, as well as editing of the remaining writeup of Task 2. (roughly 1/3 of total work)<br>



```{r}
library(VIM)
library(naniar)
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)

```


# Task 1. Comparison of complete case analysis, stochastic regression imputation and predictive mean matching






















# Task 2. Analysis of NHANES data regarding risk factors of cardiovascular disease

## 1.


```{r}
## Assignment IDA 25/26 reading in NHANES data

# set the working directory to the location where the data file is

dat<-dget(file="NHANESvIDA.rda")
str(dat) # concise info on size and variables
dat$id <- NULL # delete observation id as not required
head(dat) # first six observations
summary(dat) # summary for all variables 

```

## 2.


```{r}
# aggregation plot

v <- aggr(dat, plot=FALSE)
plot(v, prop=FALSE, numbers=TRUE)

```


```{r}
# missing stratified by categorical variable


gg_miss_var(dat %>% select(-c(marital, ethnic, educ, income, num_mvpa_bouts, cpm, age)), facet=gender, show_pct=TRUE) + ggtitle("Gender") + theme(
    plot.title = element_text(size = 16, face = "bold",  hjust = 0.5))
gg_miss_var(dat %>% select(-c(gender, marital, educ, income, num_mvpa_bouts, cpm, age)), facet=ethnic, show_pct = TRUE)  + ggtitle("Ethnicity") + theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
gg_miss_var(dat  %>% select(-c(gender, marital, ethnic, income, num_mvpa_bouts, cpm, age)), facet=educ, show_pct=TRUE)  + ggtitle("Education") + theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
gg_miss_var(dat  %>% select(-c(gender, educ, ethnic, income, num_mvpa_bouts, cpm)), facet=marital, show_pct=TRUE)  + ggtitle("Marital Status") + theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
dat_no_income_na <- dat %>% filter(!is.na(income))
gg_miss_var(dat_no_income_na  %>% select(-c(gender, marital, educ, ethnic, num_mvpa_bouts, cpm, age)), facet = income, show_pct = TRUE)  + ggtitle("Income") + theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))


```

```{r}
# get numerical summary of missingness percentages by categorical variable

cat_vars <- c("gender", "ethnic", "educ", "marital", "income")
vars_to_check <- setdiff(colnames(dat), cat_vars)
vars_with_missing <- vars_to_check[colSums(is.na(dat[vars_to_check])) > 0]
missing_by_value <- lapply(cat_vars, function(cat) {
  dat %>%
    group_by(across(all_of(cat))) %>%
    summarise(
      across(all_of(vars_with_missing), ~ mean(is.na(.)) * 100),
      .groups = "drop"
    ) %>%
    rename(value = all_of(cat)) %>%
    mutate(categorical_var = cat) %>%
    select(categorical_var, value, everything())
})
missing_by_value <- bind_rows(missing_by_value)

missing_by_value

```


```{r}
# margin plots (predictors against cholesterol)

marginplot(dat[c(10, 7)],
           main = "Height vs. Cholesterol",
           xlab = "Height (cm)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

marginplot(dat[c(10, 8)],
           main = "BMI vs. Cholesterol",
           xlab = "Body mass index (kg/m^2)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

marginplot(dat[c(10, 9)],
           main = "Waist vs. Cholesterol",
           xlab = "Waist circumference (cm)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

marginplot(dat[c(10, 11)],
           main = "Apoliprotein B vs. Cholesterol",
           xlab = "Apoliprotein (B) (mg/dL)",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

marginplot(dat[c(10, 12)],
           main = "Mean counts vs. Cholesterol",
           xlab = "Means counts per day",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

marginplot(dat[c(10, 13)],
           main = "Mean MVPA bouts vs. Cholesterol",
           xlab = "Mean MVPA bouts per day",
           ylab = "LDL-Cholesterol (mg/dL)",
           cex.main = 1.2, font.main = 2)

```


## 3.

```{r}
# univariate summaries of cholesterol

# numerical summary
print(summary(dat$cholesterol))

non_na_cholesterol <- dat[!is.na(dat$cholesterol), ]

par(mfrow = c(1, 2))

# box plot 
boxplot(non_na_cholesterol$cholesterol,
        main = "Boxplot of Cholesterol",
        ylab = "LDL-Cholesterol (mg/dL)",
        col = "lightblue",
        border = "steelblue")

# density plot
dens <- density(non_na_cholesterol$cholesterol)
plot(dens,
     main = "Density of Cholesterol",
     xlab = "LDL-Cholesterol (mg/dL)",
     col = "darkorange",
     lwd = 2)
polygon(dens, col = "moccasin", border = "darkorange")

par(mfrow = c(1, 1))

```


```{r}
# bivariate summaries

# box plots of cholesterol by each categorical variable
non_na_cholesterol <- non_na_cholesterol %>%
  filter(!is.na(income))

dat_long <- non_na_cholesterol %>%
  pivot_longer(
    cols = cat_vars,
    names_to = "variable",
    values_to = "category"
  )

ggplot(dat_long, aes(x = category, y = cholesterol)) +
  geom_boxplot(fill = "skyblue") +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  labs(
    x = "",
    y = "LDL-Cholesterol (mg/dL)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


# scatter plots of cholesterol by each numeric variable
dat_long_num <- non_na_cholesterol %>%
  pivot_longer(
    cols = c(age, height, bmi, waist, cpm, num_mvpa_bouts),
    names_to = "variable",
    values_to = "value"
  )

ggplot(dat_long_num, aes(x = value, y = cholesterol)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  labs(
    x = "Predictor Value",
    y = "LDL-Cholesterol (mg/dL)"
  )

# pearson correlation matrix plot
num_data <- dat %>%
  select(cholesterol, age, height, bmi, waist, cpm, num_mvpa_bouts) %>%
  na.omit()
cor_mat <- cor(num_data, method = "pearson")
corrplot(cor_mat,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         number.cex = 0.7)


```


## 4.




## 5.











